 Casting multiple shadows: high-dimensional interactive data visualisation with tours and embeddings

### 2020-04-20 paper.Rmd
update manuscript

There has been a rapid uptake in the use of non-linear dimensionality reduction
(NLDR) methods such as t-distributed stochastic neighbour embedding (t-SNE) in
the natural sciences as part of cluster orientation and dimension reduction
workflows. The appropriate use of these methods is made difficult by their
complex parameterisations and the multitude of decisions required to balance
the preservation of local and global structure in the resulting visualisation.
We present a visual analytics framework for the pragmatic usage of NLDR methods
by combining them with a technique called the tour. A tour is a sequence of
interpolated linear projections of multivariate data onto a lower dimensional
space. The sequence is displayed as a dynamic visualisation, allowing a user to
see the shadows the high-dimensional data casts in a lower dimensional view. By
linking the tour to a view obtained from an NLDR method, we can preserve global
structure and through user interactions like spatial linked brushing observe
where the NLDR view may be misleading. We display several case studies from
single cell genomics, that shows our approach is useful for cluster orientation
tasks.  The implementation of our framework is available as an R package called
`liminal` available at https://github.com/sa-lee/liminal.

@@

@@

(ref:fake-trees) Example from PHATE

```{r fake-trees, echo = FALSE, out.width = "50%", fig.cap="(ref:fake-trees)"}
knitr::include_graphics(here::here("img", "tree.png"))
```

### Case Study 2: Clustering PBMC 10x single cell RNA-seq data

>>>>

There has been a rapid uptake in the use of non-linear dimensionality reduction
(NLDR) methods such as t-distributed stochastic neighbour embedding (t-SNE) in
the natural sciences as part of cluster orientation and dimension reduction
workflows. The appropriate use of these methods is made difficult by their
complex parameterisations and the multitude of decisions required to balance
the preservation of local and global structure in the resulting visualisation.
We present a visual analytics framework for the pragmatic usage of NLDR methods
by combining them with a technique called the tour. A tour is a sequence of
interpolated linear projections of multivariate data onto a lower dimensional
space. The sequence is displayed as a dynamic visualisation, allowing a user to
see the shadows the high-dimensional data casts in a lower dimensional view. By
linking the tour to a view obtained from an NLDR method, we can preserve global
structure and through user interactions like spatial linked brushing observe
where the NLDR view may be misleading. We display several case studies from
single cell transciptomics and image processing, that shows our approach is 
useful for cluster orientation tasks.  The implementation of our framework 
is available as an R package called
`liminal` available at https://github.com/sa-lee/liminal.

@@

The next section steps through three use cases of our approach 
using simulated and real data. In the first case study, we step through
data that lies on tree structure (similar to trajectories in single
cell transcriptomics) to that highlights an iterative approach to modifying
t-SNE views based on what we learn in the high-dimensional space. We show
how through simple linked brushing, we can see when the assumptions
of continuity of the embedding map are broken. In the second case study,
we explore a well-clustered single cell transcriptomics data of mice retina...
Finally, in the last case study we explore a subset of the QuickDraw dataset,
we use `liminal` to identify instances of images from different labels that
appear similiar.

@@

(ref:fake-trees) Example high-dimensional tree shaped data, $n = 3000$ and $p = 100$. The data lies 2-d tree like structure consisting of ten branches. This data is available in the `phateR` package and is simulated via diffusion-limited aggregation (a random walk along the branches of the tree) with Gaussian noise added [@Moon2019-ce].

```{r fake-trees, echo = FALSE, out.width = "50%", fig.align = "center", fig.cap="(ref:fake-trees)"}
knitr::include_graphics(here::here("img", "tree.png"))
```

### Case Study 2: Clustering single cell RNA-seq data
